%TC:group tabular 1 1
%TC:group table 1 1

\chapter{Conclusion}

\paragraph{} The project set out to explore the hypothesis that spreadsheets would provide a productive medium for musical expression. Excello is a notation and corresponding program for musical playback integrated within Microsoft's Excel. By abstracting time away from the axes of the grid, the existing functionality of Excel remains highly useful. Having satisfied the initial success criteria for the program, development continued, carrying out participatory design with 21 users. As a result of this, many additional features, beyond the initial scope of the project, were implemented all of which have been shown to significantly improve the interface. With respect to CDN, reasonable closeness of mapping, high consistency, high secondary notation, low viscosity and high visibility were all achieved as desired. Quantitatively, Excello is able to express any  MIDI music, and a converter was built to translate existing corpora of MIDI files to CSV files for Excello. The converter included two additional compression mechanisms, which still represent all musical information under certain, but common, conditions.

\paragraph{} During development, I submitted part of my code as an improvement to the open-source library Parenthesis. This was merged and has been published. The package has over 20,000 weekly npm downloads.

\paragraph{} Excello freely provides a simple, but powerful program for musical composition to the hundreds of millions of users already familiar with the spreadsheet interface.
